Immutable Source-Mapped Abstract Syntax Tree:\ \, \\ A Design Pattern for Refactoring Engine APIs

### 2017-10-04 paper.tex
final changes

\begin{itemize}
    \item \textit{The API should describe the program's structure at a specific 
        point in time, either before or after the a refactoring transformation.} 
        It should never be inconsistent and provide some information that is
        up-to-date with the latest transformation while providing some
        informations that isn't.
\end{itemize}
\clearpage
\begin{itemize}
    \item \textit{The API should be robust in the presence of attempted refactorings
        that produce invalid code.} The refactoring should never leave a 
        client with code that doesn't compile. 
    \item \textit{The API should be implemented in a way that allows easy analysis of
        the syntactic effects of any given change.} This implementation should
        be memory and time efficient, allowing clients to write refactorings
        that meet the performance requirements of an interactive tool.
    \item \textit{The API should be robust in the presence of "malicious" clients.}
        its implementation should not allow clients to manipulate or 
        invalidate internal data structures. This requires an understanding of
        possible states of those data structures. In particular, an AST used
        as an internal data structure might be exposed differently to clients
        via the API. 
\end{itemize}

%TODO work on this preprocessor stuff some more
%Maybe move to a different place?
\subsection{Preprocessors}
Preprocessors (such as the C preprocessor and M4), don't interfere with the ASTs
used by compilers. In most cases, the preprocessing is performed first, and the
output thereof is used to construct the AST. The lexer/parser requires no 
knowledge of the fact that the text was preprocessed.

Refactorings, however, are complicated by the C preprocessor. The grammar of 
preprocessor directives doesn't take into account C syntax, and can't be easily
represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
\cite{garrido05program}. The compiler's method of preprocessing before parsing
can't be used, because the unpreprocessed text is what must be edited. The 
refactoring too must be able to parse, analyze, and transform code \textit{with embedded preprocessor directives.}

The topic of refactoring preprocessed code will not be treated in detail here,
but suffice it to say that a compiler's AST will almost certainly be inadequate.
For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.

\section{Solution}

@@

Now, consider what happens when a user selects text in the text editor.  This
can be represented as an offset-length pair.  When source location information
is available for every AST node, it becomes fairly easy to determine what
syntactic the text selection corresponds to: simply
walk the tree, searching for a node whose source location aligns with the
user's selection.

@@

>>>>

\begin{itemize}
    \item \textit{The API should describe the program's structure at a specific 
        point in time, either before or after the a refactoring transformation.} 
        It should never be inconsistent and provide some information that is
        up-to-date with the latest transformation while providing some
        information that isn't.
\end{itemize}
\clearpage
\begin{itemize}
    \item \textit{The API should be robust in the presence of attempted refactorings
        that produce invalid code.} The refactoring should never leave a 
        client with code that doesn't compile. 
    \item \textit{The API should be implemented in a way that allows easy analysis of
        the syntactic effects of any given change.} This implementation should
        be memory and time efficient, allowing clients to write refactorings
        that meet the performance requirements of an interactive tool.
    \item \textit{The API should be robust in the presence of "malicious" clients.}
        Its implementation should not allow clients to manipulate or 
        invalidate internal data structures. This requires an understanding of
        possible states of those data structures. In particular, an AST used
        as an internal data structure might be exposed differently to clients
        via the API. 
\end{itemize}

%TODO work on this preprocessor stuff some more
%Maybe move to a different place?
\section{Solution}

@@

@@

Now, consider what happens when a user selects text in the text editor.  This
can be represented as an offset-length pair.  When source location information
is available for every AST node, it becomes fairly easy to determine what
syntactic structure the text selection corresponds to: simply
walk the tree, searching for a node whose source location aligns with the
user's selection.

This mapping, however, is complicated by the C preprocessor. The grammar of 
preprocessor directives doesn't take into account C syntax, and can't be easily
represented in the C AST. This is the subject of Garrido's entire Ph.D.~thesis~%
\cite{garrido05program}. The topic of refactoring preprocessed code will not be treated in detail here,
but suffice it to say that a compiler's AST will almost certainly be inadequate.
For more details, see Garrido~\cite{garrido2013} and Overbey et al.~\cite{wrt2009}.
